---
title: "02_homework1_part1"
author: "Randy"
date: "3/9/2021"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE, cache=TRUE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```


```{r include=FALSE}
library("mgcv")
library("tidymodels")
library("tidyverse")
library("here")
library("janitor")
library("knitr")
library("tinytex")
library("bookdown")
```

# question1
## 1a

$$ 
y_i = f(x_i) + \epsilon_i 
$$
$$
\epsilon_i \sim Normal(0, \sigma_{\epsilon}^2) 
$$

$$
f(x_i) = \sum_{k=1}^K\xi_k \ \phi_k(x)
$$

$$ 
PENSSE_\lambda = \sum_{i=1}^N (y_i - f(x_i))^2 + \lambda \int{f''(x)^2}dx
= \sum_{i=1}^N (y_i - f(x_i))^2 + \pmb{\xi^{\top} S \  \xi }
$$

The continuous function of f(x) must have at least two derivatives. 

$$
\text{let } \delta_j(x) = \phi''_j(x) \\
f(x_i) = \sum_{k=1}^K\xi_k \ \phi_k(x_i) \  =\  \pmb{\xi^{\top}\phi(x)} \ =\  \pmb{\phi(x)^{\top} \xi} \\
$$

$$
f''(x_i) = \pmb{\xi^{\top}\delta(x)} = \pmb{\delta(x)^{\top} \xi}  
$$
$$ 
\lambda \int{f''(x)^2}dx = \lambda \ \pmb{\xi^{\top}} 
\bigg( \int \pmb{\delta(t) \delta^{\top}(t)} dt\ \bigg) \ \pmb{\xi^{\top}} 
$$
$$
let \ \ \ \pmb{S} = \int \pmb{\delta(t) \ \delta^{\top}(t)} dt\ 
$$
$$
PENSSE_\lambda = \sum_{i=1}^N (y_i - f(x_i))^2 + \pmb{\xi^{\top} S \  \xi }
$$


## 1b

$$
f(x_i) = \sum_{k=0}^3\xi_k \ x^k \ = \xi_0 + \xi_1x + \xi_2x^2 + \xi_3x^3 \\
f'(x) = \xi_1 + 2\xi_2x + 3\xi_3x^2 \\
f''(x) = 2\xi_2 + 6\xi_3x = 
\pmb{\xi^{\top}\delta(x)}
\\
\pmb {\delta(x)} = [0, 0, 2, 6x]
\\
\pmb{S} = \int \pmb{\delta(t) \ \delta^{\top}(t)} dt
$$

$$
\pmb {\delta(t) \delta^{\top}(t)}=
\begin{bmatrix}
0, 0, \ \ \ \ 0,\ \ \ \ \ \ 0\\
0, 0, \ \ \ \ 0,\ \ \ \ \ \ 0\\
0, 0, \ \ \ \ 4, \ \ 12x\\
0, 0,     12x, 36x^2
\end{bmatrix}
\\
$$

$$
\pmb{S} = \int^{10}_{0} \pmb{\delta(t) \ \delta^{\top}(t)} dt =
\begin{bmatrix}
0, 0, \ \ \ \ \ \ \ 0,\ \ \ \ \ \ \ \ \ 0\\
0, 0, \ \ \ \ \ \ \ 0,\ \ \ \ \ \ \ \ \ 0\\
0, 0,  4x|_0^{10}, \ \ 6x^2|_0^{10}\\
0, 0,     6x^2|_0^{10}, 12x^3|_0^{10} 
\end{bmatrix} =
\begin{bmatrix}
0, 0, \ \ \ \ 0,\ \ \ \ \ \ \ \ 0\\
0, 0, \ \ \ \ 0,\ \ \ \ \ \ \ \ 0\\
0, 0, \ \  40, \ \ \ \ 600\\
0, 0,     600, 12000 
\end{bmatrix}
$$


## 1c
$$
PENSSE_\lambda = \sum_{i=1}^N (y_i - f(x_i))^2 + \lambda\ \pmb{\xi^{\top} S \  \xi } \\
= \| \pmb{y} - \pmb{\Phi \xi} \|^2 + \
\lambda\ \pmb{\xi^{\top} S \xi } 
\\
= (\pmb{y} - \pmb{\Phi \xi})^{\top}(\pmb{y} - \pmb{\Phi \xi}) +
\lambda\ \pmb{\xi^{\top} S \  \xi }
\\
= \pmb{y}^{\top} \pmb{y} - 2\pmb{\xi}^{\top}\pmb{\Phi}^{\top}\pmb{y} + \pmb{\xi}^{\top}(\pmb{\Phi^{\top}\Phi)} \pmb{\xi} +
\lambda\ \pmb{\xi^{\top} S \  \xi } \\
= \pmb{y}^{\top} \pmb{y} - 2\pmb{\xi}^{\top}\pmb{\Phi}^{\top}\pmb{y} + \pmb{\xi}^{\top}(\pmb{\Phi^{\top}\Phi + \lambda S)} \pmb{\xi}
$$

$$
\nabla_{\pmb {\xi}} PENSSE_{\lambda}  = 
-2\pmb{\Phi^{\top} y} + 2(\pmb{\Phi^{\top}\Phi} + 
\lambda \pmb {S})\pmb{\xi} \stackrel{set}{=} 0
$$

$$
2(\pmb {\Phi^{\top} \Phi} + 
\lambda \pmb {S})\pmb{\xi} = 2\pmb{\Phi^{\top} y}
$$

$$
\hat {\pmb{\xi}} = (\pmb{\Phi^{\top}\Phi + 
\lambda S})^{-1} \pmb{\Phi^{\top} y}
$$

$$
\nabla^2_{\pmb {\xi}} PENSSE_{\lambda} = 
2(\pmb{\Phi^{\top}\Phi + \lambda S}) > 0
$$

$$
\hat{\xi} \ minimized\  the\  PENSSE_{\lambda}
$$

$$
Cov(\pmb y) = \sigma^2 \pmb {\mathcal I} 
$$

$$
Cov(\hat \xi) = (\pmb{\Phi^{\top}\Phi + \lambda S})^{-1} \pmb{\Phi^{\top} }
                Cov(y)
                \pmb{\Phi}(\pmb{\Phi^{\top}\Phi} + \lambda \pmb {S})^{-1}
                \\
                = \sigma^2(\pmb{\Phi^{\top}\Phi} + \lambda \pmb {S})^{-2}
$$

## 1d
$$
\hat {\pmb{\xi}} = (\pmb{\Phi^{\top}\Phi + 
\lambda S})^{-1} \pmb{\Phi^{\top} y}
\\
\pmb H = \pmb{\Phi}(\pmb{\Phi^{\top}\Phi + 
\lambda S})^{-1} \pmb{\Phi^{\top} }
\\
\hat{f}(x|\lambda) = \pmb{\Phi \hat \xi} = \pmb {H y}
\\
Var[\hat{f}(x|\lambda)] = Var[\pmb{Hy}] 
= \pmb{H} Var[\pmb y] \pmb{H^\top} 
\\
(\text{H is the hat matrix not a vector.}）
\\
$$

## 1e

now we define the equation holds for the augmented matrices:
$$
\begin{Vmatrix}
\begin{bmatrix}
y\\ \pmb {\mathcal a} 
\end{bmatrix} -
\begin{bmatrix}
\pmb{\Phi} \\
\pmb {\mathcal B} 
\end{bmatrix} \pmb {\xi} 
\end{Vmatrix} ^ 2 =
\|y - \Phi \pmb {\xi} \|^2 + 
\| \mathcal a - \mathcal B \pmb {\xi} \|^2 
\\
= \|y - \Phi \pmb {\xi}\|^2 + 
\lambda \pmb {\xi^{\top} S \pmb {\xi}}
\\
= \|y - \Phi \pmb{\xi}\|^2 + 
\lambda \pmb{\xi^{\top} D^{\top} D \xi}
$$

$$
\| \pmb {\mathcal a} - \pmb{\mathcal B \xi} \|^2 
\\
= (\pmb a - \pmb {\mathcal B \xi})^{\top}(\pmb {a- \mathcal B\xi})
\\
= \pmb{ a^{\top}a - 2\xi^{\top}\mathcal B ^{\top}a + \xi^{\top}\mathcal B^{\top} \mathcal B \xi }
\\
= (\pmb{ a^{\top} - 2\xi^{\top}B^{\top})a +  \xi^{\top}\mathcal B^{\top} \mathcal B \xi}
\\ 
= \lambda \pmb {\xi^{\top}D^{\top}D  \xi}
$$

$$
\text{To make the the equation work:} \\ 
\lambda \pmb \xi^{\top} \pmb D^{\top}\pmb D \xi
= \lambda \pmb {\xi^{\top} S \xi}
\\
(\pmb{ a^{\top} - 2\xi^{\top} \mathcal B^{\top})a} =0, \forall \ \pmb{\mathcal B} ,  \pmb{\xi} 
\\
Then, \ \pmb a =  \pmb 0, \ and \ \pmb{\mathcal B} = \lambda^{1/2} \pmb D
$$

## 1f

The parameters depend on each other, so it is computational challenging. For a likelihood function depends on more than one parameters, the profile likelihood should be applied. 


# question2
```{r message=FALSE, warning=FALSE}
set.seed(5520)
N <- 100
sigma = 0.5
f <- function(x) cos(2 * pi * x)
x <- runif(N, min = -1, max = 1)
y <- f(x) + rnorm(N, mean = 0, sd = sigma)

## choose the lambda
## choose a very large value of K
## cross validation
sm_cr <-
  ## directly from the lecture2 notes
  mgcv::smoothCon(
    s(x, ## a list of variables as covariates
      ## this smooth is a function of
      bs = "cr",
      ## bs for penalized smoothing basis
      ## cr for cubic regression spline
      k = 25),
    data = data.frame(x = x))
## get the "S" matrix
s_matrix <- sm_cr[[1]]$S[[1]]

nlambda <- 1000
loglambda <- seq(-10, 20, len = nlambda)

phi <- sm_cr[[1]]$X
## get the spline basis matrix
## the design matrix
# phi_phi <- t(Phi) %*% Phi
# phi_prod <- crossprod(Phi)
# identical(phi_phi, phi_prod)

mse_cv <- yhat <- yse <- rep(NA, nlambda)
for (i in seq_along(1:nlambda)) {
  ## do the cross-validation
  ## to get the inverse matrix
  ## the standard R function with solve()
  solution <- solve(crossprod(phi) + 
                      exp(loglambda[i]) * s_matrix)
  ## A is the influence hat matrix
  h_matrix <- phi %*% solution %*% t(phi)
  yhat[i] <- h_matrix %*% y
  yse[i] <- norm(h_matrix) %*% sd(y)
  ## mean square error cross validation
  mse_cv[i] <- mean((yhat[i] - y)^2 / 
                      (1 - diag(h_matrix))^2)
}

loglambda[which.min(mse_cv)]
lambda_min <- loglambda[which.min(mse_cv)] %>% exp()
lambda_min


```

write it up in a function
```{r}
#' @description function to get the optimal penalty parameter
#' @para phi the smooth basis from the original data
#' @para s_matrix the penalty matrix 
#' @para y outcome vector
#' @para nl the number of lambda in the trial
#' @para llmin the minimum of the log lambda 
#' @para llmax the maximum of the log lambda
#' @para tolerance for the inverse function to avoid singularity
#' @example see below the function
get_lambda <- function(phi, 
                      s_matrix, 
                      y,
                      nl = 1000, 
                      llmin = -10, 
                      llmax = 10,
                      tolerance = 1e-26) {

  loglambda <- seq(llmin, llmax, len = nl)
  
  mse_cv <- rep(NA, nlambda)
  for (i in 1:nlambda) {
    ## do the cross-validation
    ## to get the inverse matrix
    ## the standard R function with solve()
    solution <- solve(crossprod(phi) + 
                      exp(loglambda[i]) * s_matrix,
                      tol = tolerance)
    ## A is the influence hat matrix
    h_matrix <- phi %*% solution %*% t(phi)
    yhat <- h_matrix %*% y
    ## mean square error cross validation
    mse_cv[i] <- mean((yhat - y)^2 / (1 - diag(h_matrix))^2)
    }
  lambda_min <- loglambda[which.min(mse_cv)] %>% exp()
  return(lambda_min)
}
```


```{r}
get_xi <- 
  function(y, phi, s_matrix, lambda){
    solution <- solve(crossprod(phi) + lambda * s_matrix)
    xi <- solution %*% t(phi) %*% y %>%
      as.data.frame() %>%
      select("est_xi" = 1)
    var_xi <- solution %*% t(phi) %*% 
              (cov(y, y) * diag(length(y))) %*% 
              phi %*% solution 
    
    se_xi <- var_xi %>%
      diag() %>%
      sqrt()
    
    se_yfit <- phi %*% var_xi %*% t(phi) %>%
      diag() %>%
      sqrt()
    # Tue Mar 16 14:54:01 2021 ------------------------------
    
    xi_table <- cbind(xi, se_xi, se_yfit) %>%
      as.data.frame() 
    
    h_matrix <- phi %*% solution %*% t(phi)
    yhat <- h_matrix %*% y
    yse <- norm(h_matrix) %*% cov(y, y) %>%
      diag() %>% sqrt()
    ## mean square error cross validation
    attr(xi_table, "yhat") <- yhat
    attr(xi_table, "yse") <- yse
    
    return(xi_table)
  }

xi <- get_xi(y, phi, s_matrix, lambda_min)
xi %>% 
  select(est_xi, se_xi) %>%
  head(10) %>%
  mutate(var_xi = se_xi^2) 

## this is for question3b,
## just please do not do this again....
yhat <- get_xi(y, phi, s_matrix, lambda_min) %>%
  attr("yhat")
yse <- get_xi(y, phi, s_matrix, lambda_min) %>%
  attr("yse") 

ypred_2 <-
  cbind(x, yhat, yse) %>%
  data.frame() %>%
  mutate(lower = yhat - 2 * yse,
         upper = yhat + 2 * yse,
         question = "question2") %>%
  select(yhat = 2, everything())
```

# question3
## 3a

$$
\epsilon_i \sim Normal(0, \ \sigma_{\epsilon}^2),
\\
\epsilon_i = y_i - f(x_i) = y_i - \sum_{k=1}^K\xi_k \phi_k(x_i),
\\
\pmb V \equiv Cov (\pmb {\epsilon}) = \sigma_{\epsilon}^2 \pmb {\mathcal I}
\\
l(\pmb \xi, \sigma^2_{\epsilon}|\pmb y, \pmb \Phi( \pmb x)) = 
-\frac {nlog(2\pi \sigma_{\epsilon}^2)} {2}-\sum_{i=1}^n (y_i - \sum_{k=1}^K\xi_k \phi_k(x_i))^2 /2\sigma^2_{\epsilon} 
\\
= -\frac {nlog(2\pi) + log|\pmb V|+ (\pmb y- \pmb \Phi \pmb{\xi})^{\top} \pmb V ^{-1} (\pmb {y- \Phi\xi})} {2}
$$

$$
pl(\pmb \xi, \sigma^2_{\epsilon}|\pmb y, \pmb \Phi( \pmb x), \lambda) = 
l(\pmb \xi, \sigma^2_{\epsilon}|\pmb y, \pmb \Phi( \pmb x)) - 0.5\lambda \pmb {\xi^{\top} S \xi}
\\
= - \frac {nlog(2\pi) + log|V|+ 
(\pmb y- \pmb \Phi \pmb{\xi})^{\top}V^{-1} (\pmb {y- \Phi \xi})} {2} - 0.5\lambda \pmb {\xi^{\top} S \xi}
\\
\nabla_{\pmb \xi} \ \mathcal pl(\pmb \xi, \sigma^2_{\epsilon}|\pmb y,\pmb\Phi ( \pmb x), \lambda) \stackrel{set}{=} 0
\\
\pmb{\Phi^{\top}V^{-1}(y - \Phi\xi)} - \lambda \pmb  {S \xi} \stackrel{set}{=} 0
\\
\pmb{\Phi^{\top}V^{-1}\Phi\xi} + \lambda \pmb  {S \xi} = \pmb {\Phi^{\top}V^{-1}y}
\\
\pmb {\hat \xi} = \pmb{(\Phi^{\top}V^{-1}\Phi} + \lambda \pmb {S})^{-1} \pmb{\Phi^{\top}V^{-1}y}
\\
=\pmb{(\Phi^{\top} \Phi} + \sigma_{\epsilon}^2  \lambda \pmb {S})^{-1} \pmb{\Phi^{\top} y}
$$

$$
\\
\nabla^2_{\pmb \xi} \ \mathcal pl(\pmb \xi, \sigma^2_{\epsilon}|\pmb y,\pmb\Phi(\pmb{x}), \lambda) = 
- \pmb {\Phi^{\top}V^{-1}\Phi} - \lambda \pmb{S}= - \pmb {\mathcal I_n(\hat \xi)}
\\
\pmb {\hat \xi} \sim Normal(\pmb \xi, \pmb {\mathcal I_n(\hat \xi)}^{-1})
\\
Cov(\pmb{\hat\xi}) = (\pmb {\Phi^{\top}V^{-1}\Phi} + \lambda \pmb{S})^{-1}
\\
Var[\pmb{\hat\xi}] = diag\big((\pmb {\Phi^{\top}V^{-1}\Phi} + \lambda \pmb{S})^{-1}\big)
\\
\hat {Var}[\hat\xi_k]= \sqrt [] {diagonal \ k^{th}\ of\
(\pmb {\Phi^{\top}V^{-1}\Phi} + \lambda \pmb{S})^{-1} }
$$

1) by default the optim() function performs minimization 
(as opposed to maximization)
2) optimize on the log-likelihood scale
3) using the optim() function, 
set the argument hessian = TRUE to obtain the Hessian matrix

```{r}


```

```{r}
set.seed(555)
xi_sim <- 
  optim(rep(500, 25),
        fn = get_maxpl,
        ## use quasi newton
        method = "BFGS", 
        ## keep the hessian
        hessian = TRUE,
        lambda = lambda_min,
        sigma = sigma,
        X = x,
        y = y)
est_xi3 <- xi_sim$par
## variance is the inverse of the hessian
se_xi3 <- solve(xi_sim$hessian) %>%
  diag() %>%
  sqrt()

# get_maxpl(est_xi3, lambda_min, sigma, x, y) %>%
#   attr("xi_hat") %>%
#   unlist() 
# get_maxpl(est_xi3, lambda_min, sigma, x, y) %>%
#   attr("xi_se") 
## Sat Mar 13 16:56:47 2021 ------------------------------
## xi_hat use 2 times of the penalty terms
# [1,]  0.7058491 # [2,]  0.6602813 # [3,]  0.1364618 # [4,] -0.4658330
# [5,] -0.5886410 # [6,] -0.7478183 # [7,] -0.7407876 # [8,] -0.8627927
# [9,] -0.9326031 # [10,] -0.6079508 # [11,] -0.1056475 # [12,]  0.5074190
# [13,]  0.9322000 # [14,]  0.8950186 # [15,]  0.5166366 # [16,]  0.1626429
# [17,] -0.3897756 # [18,] -0.7906138 # [19,] -0.7254311 # [20,] -0.8621898
# [21,] -0.8276467 # [22,] -0.1506179 # [23,]  0.6910463 # [24,]  1.1297209
# [25,]  1.8903166
# Sat Mar 13 16:57:22 2021 ------------------------------
```

## 3b
```{r}
# get_maxpl(est_xi3, lambda_min, sigma, x, y) %>%
#   attr("h") 
yhat <- get_maxpl(est_xi3, lambda_min, sigma, x, y) %>%
  attr("yhat")
yse <- get_maxpl(est_xi3, lambda_min, sigma, x, y) %>%
  attr("yse") 

ypred_3 <-
  cbind(x, yhat, yse) %>%
  data.frame() %>%
  mutate(lower = yhat - 2 * yse,
         upper = yhat + 2 * yse,
         question = "question3") %>%
  select(yhat = 2, everything())

rbind(ypred_2, ypred_3) %>%
  ggplot() +
  geom_line(aes(x = x, y = yhat, 
                group = question, color = question) ) +
  geom_line(aes(x = x, y = upper, group = question, color = question),
            linetype = "dashed") +
  geom_line(aes(x = x, y = lower, group = question, color = question),
            linetype = "dashed") + 
  theme_classic()

```
The estimated trajectory and confidence interval from two different methods are 
exactly the same as each other. The cross-validation is consistent with penalized maximum likelihood estimation 


# question4
## 4a

The ploy() did not include an intercept term in matrix;
the extra column and row of 1 can be added manually (but not included in this homework).
```{r}
set.seed(19870)

N <- 1000
f <- function(x, a) a * x^2
x1 <- runif(N, 0, 1)
y1 <- f(x1, a = 5) + rnorm(N, sd = 0.5)
x2 <- rexp(N, rate = 1 / 100)
y2 <- f(x2, a = 1e-05) + rnorm(N, sd = 0.5)
K = 5
data4 <- data.frame(y1, y2, x1, x2)

phi_1 <- poly(x1, k = K, raw = TRUE)
phi_2 <- poly(x2, k = K, raw = TRUE)

## build the penalty matrix S_x
S_x1 <- matrix(rep(NA, K * K), nrow = K, ncol = K)
for (i in 1:K) {
  for (j in 1:K) {
    ## the derivative for the equation
    fun <- function(t) i * (i - 1) * t^(i + j - 4) * j * (j - 1)
    ## integrate over the range of x
    a <- integrate(fun, lower = min(x1), upper = max(x1))
    S_x1[j, i] <- a$value
  }
}
S_x2 <- matrix(rep(NA, K * K), nrow = K, ncol = K)
for (i in 1:K) {
  for (j in 1:K) {
    ## the derivative for the equation
    fun <- function(t) i * (i - 1) * t^(i + j - 4) * j * (j - 1)
    ## integrate over the range of x
    a <- integrate(fun, lower = min(x2), upper = max(x2))
    S_x2[j, i] <- a$value
  }
}

S_x1
S_x2
```

## 4b 
unpenalized least squares

```{r}
xi_4b_hat1 <- solve(t(phi_1) %*% phi_1) %*% t(phi_1) %*% y1
xi_4b_hat2 <- solve(t(phi_2) %*% phi_2, tol = 1e-26) %*% t(phi_2) %*% y2
xi_4b_gin2 <- MASS::ginv(t(phi_2) %*% phi_2) %*% t(phi_2) %*% y2

xi_4b_cov1 <- 
  solve(t(phi_1) %*% phi_1) %*% t(phi_1) %*% 
  (cov(y1, y1) * diag(length(y1))) %*% 
  phi_1 %*% solve(t(phi_1) %*% phi_1) 
xi_4b_cov2 <- 
  solve(t(phi_2) %*% phi_2, tol = 1e-26) %*% t(phi_2) %*% 
  (cov(y2, y2) * diag(length(y2))) %*% 
  phi_2 %*% solve(t(phi_2) %*% phi_2, tol = 1e-26) 

xi_4b_hat1
xi_4b_hat2
xi_4b_gin2
# Fri Mar 26 21:43:45 2021 ------------------------------
gam4_reml1 <- gam(y1 ~ phi_1 - 1,
              method = "REML",
              family = gaussian(),
              data = data4)
gam4_reml2 <- gam(y2 ~ phi_2 - 1,
              method = "REML",
              family = gaussian(),
              data = data4)
coef(gam4_reml1)
coef(gam4_reml2)
```

## 4c 
penalized mean error square

```{r message=FALSE, warning=FALSE}
# lambda1 <- get_lambda(phi = phi_1, s_matrix = S_x1, y = y1)
## reset the loglambda upper bound to 50,
## increasing tolerance for inverse function to 1e-100 
# lambda2 <- get_lambda(phi = phi_2, s_matrix = S_x2, y = y2, llmax = 50, tol = 1e-100)
lambda1 <- 0.003944
lambda2 <- 2114998
xi_4c_hat1 <- solve(t(phi_1) %*% phi_1 + 0.25 * lambda1 * S_x1) %*% t(phi_1) %*% y1
xi_4c_hat2 <- solve(t(phi_2) %*% phi_2 + 0.25 * lambda2 * S_x2, tol = 1e-26) %*% t(phi_2) %*% y2
xi_4c_gin2 <- MASS::ginv(t(phi_2) %*% phi_2 + 0.25 * lambda2 * S_x2) %*% t(phi_2) %*% y2

xi_4c_cov1 <- 
  solve(t(phi_1) %*% phi_1 + 0.25 * lambda1 * S_x1) %*% t(phi_1) %*% 
  (cov(y1, y1) * diag(length(y1))) %*% 
  phi_1 %*% solve(t(phi_1) %*% phi_1 + 0.25 * lambda1 * S_x1)
xi_4c_cov2 <- 
  solve(t(phi_2) %*% phi_2 + 0.25 * lambda2 * S_x2, tol = 1e-26) %*% t(phi_2) %*% 
  (cov(y2, y2) * diag(length(y2))) %*% 
  phi_2 %*% solve(t(phi_2) %*% phi_2 + 0.25 * lambda2 * S_x2, tol = 1e-26)

xi_4c_hat1
xi_4c_hat2
xi_4c_gin2

## checking with the gam or lm functions
gam4_gcv1 <- gam(y1 ~ phi_1 + 0,
              method = "GCV.Cp",
              family = gaussian(), 
              sp = -lambda1,
              data = data4)
gam4_gcv2 <- gam(y2 ~ phi_2 + 0,
              method = "GCV.Cp", 
              family = gaussian(), 
              sp = -lambda2,
              data = data4)
coef(gam4_gcv1)
coef(gam4_gcv2)
## get estimated coefficients
## from gam with the sp = -lambda
# phi_11     phi_12     phi_13     phi_14     phi_15 
# 1.258018  -7.782146  40.321256 -49.585510  20.847280 
# phi_21        phi_22        phi_23        phi_24        phi_25 
# 2.343784e-12  4.647032e-10  7.040575e-08 -1.441064e-10  9.020567e-14 
```

## 4d

```{r}
#' @description function to plot the prediction of outcomes with 95% CI
#' @para x the independent variables
#' @para y the outcomes
#' @para data the original dataset
#' @para phi the spline terms base on the independent variables
#' @para est the estimators calculated by hand
#' @para cov the variance-covariance matrix of the estimator
#' @example see below the function
get_est_plot <- function(x, y, 
                         data, phi,
                         est, cov) {
  y_hat <- phi %*% est
  y_se <- phi %*% cov %*% t(phi) %>%
    diag() %>%
    sqrt()
  
  plot <-  ggplot() +
    geom_point(aes(x, y), data = data, col = "grey", alpha = 0.5) +
    geom_line(aes(x, y_hat), col = "red") +
    geom_line(aes(x, y_hat + 1.96 * y_se), col = "brown") +
    geom_line(aes(x, y_hat - 1.96 * y_se), col = "brown") +
    theme_bw()
  return(plot)
}

get_est_plot(x1, y1, data4, phi = phi_1, est = xi_4b_hat1, cov = xi_4b_cov1)
get_est_plot(x2, y2, data4, phi = phi_2, est = xi_4b_hat2, cov = xi_4b_cov2)
get_est_plot(x1, y1, data4, phi = phi_1, est = xi_4c_hat1, cov = xi_4c_cov1)
get_est_plot(x2, y2, data4, phi = phi_2, est = xi_4c_hat2, cov = xi_4c_cov2)
```

```{r eval=FALSE, include=TRUE}
## not sure what is going on here
## there is some pathological behavior
## for the gam() and lm() function

get_gam_plot <- function(x, y, length = 1000,
                         data, model){
  xind_pred<- seq(min(x), max(x),len = 100)
  df_new <- data.frame(x = xind_pred)
  model = gam4_reml1
  predict <- predict(model,
                     newdata = df_new,
                     type = "response",
                     se.fit = TRUE)
  ## Sat Mar 27 12:20:47 2021 ------------------------------
  ## not exactly sure what is going on
  ## the broom.mixed::augment does not work here
  # broom::augment(model)
  plot_data <- data.frame(predict$fit, predict$se.fit, xind_pred) %>%
    select("pred" = 1, "se" = 2, "x" =3) %>%
    mutate(upper = pred + 1.96 * se,
           lower = pred - 1.96 * se)

  plot <- plot_data %>%
    ggplot() +
    geom_point(aes(x = x, y = y), data = data, color="grey", alpha = 0.5) +
    geom_line(aes(x = x, y = pred), color = "red") +
    geom_line(aes(x = x, y = upper), color = "brown") +
    geom_line(aes(x = x, y = lower), color = "brown") +
    theme_classic()
  plot
  return(plot)
}

get_gam_plot(x = x1, y = y1, data = data4, model = gam4_reml1)
get_gam_plot(x = x2, y = y2, data = data4, model = gam4_reml2)
get_gam_plot(x = x1, y = y1, data = data4, model = gam4_gcv1)
get_gam_plot(x = x2, y = y2, data = data4, model = gam4_gcv2)
```


## 4e
the change rate of polynomial gets large, 
and lambda gets to infinity; 
very large penalty add to the model so the model is exact linear form.

```{r}
## set lambda to a very large number
lambda3 <- 1e99
xi_4e_hat1 <- 
  solve(t(phi_1) %*% phi_1 + 0.25 * lambda3 * S_x1, tol = 1e-200) %*% t(phi_1) %*% y1
xi_4e_cov1 <- 
  solve(t(phi_1) %*% phi_1 + 0.25 * lambda3 * S_x1, tol = 1e-200) %*% t(phi_1) %*% 
  (cov(y1, y1) * diag(length(y1))) %*% 
  phi_1 %*% solve(t(phi_1) %*% phi_1 + 0.25 * lambda3 * S_x1, tol = 1e-200)
xi_4e_hat2 <- 
  solve(t(phi_2) %*% phi_2 + 0.25 * lambda3 * S_x2, tol = 1e-200) %*% t(phi_2) %*% y2
xi_4e_cov2 <- 
  solve(t(phi_2) %*% phi_2 + 0.25 * lambda3 * S_x2, tol = 1e-200) %*% t(phi_2) %*% 
  (cov(y2, y2) * diag(length(y2))) %*% 
  phi_2 %*% solve(t(phi_2) %*% phi_2 + 0.25 * lambda3 * S_x2, tol = 1e-200)

get_est_plot(x1, y1, data4, phi = phi_1, 
             est = xi_4e_hat1, cov = xi_4e_cov1)
get_est_plot(x2, y2, data4, phi = phi_2, 
             est = xi_4e_hat2, cov = xi_4e_cov2)
```

In this case only the linear term does not suffer the heavy penalty on the wiggli-ness.
so when the smooth parameter goes to infinity, the fitted model will be simply a simple linear regression.

# question5

I just realized that I mistakenly set the error standard deviance as 05, other than 0.5.
The bias and mse from the simulated data is pretty large.  
I used gam() fit the model and I just do not want to write all the thing manually again. 
```{r}
set.seed(555)

#' @description definitely should not write this functioon
#' @para N simulation numbers
#' @para K function numbers
#' @para f_list a list of K functions
#' @para sigma variance for the error term
#' @para bs spline regression form in s
#' @para method the gam() estimation methods
#' @example see below
simulate_gam <- function(N = 200, 
                         K = 3, 
                         f_list, 
                         sigma = 05,
                         bs = "cr", 
                         knot = 10,
                         method = "GCV.Cp") {
  if (K == length(f_list)) {
    X <- matrix(rnorm(N * K), N, K)
    i <- 1
    formula <- "y ~ 1"
    for (i in 1:K) {
      eta <- 1 + f_list[[i]](X[, i])
      term <- paste0("s(X", i, ", bs = '",
                     bs, "', k = ", knot, ")")
      formula <- paste0(formula, " + ", term)
    }
    ## Sun Mar 28 09:09:21 2021 ------------------------------
    ## I made the gam more complicated than required
    y <- eta + rnorm(N, sd = sigma)
    df_fit <- data.frame(X, y)
    formula <- as.formula(formula)
    fit <- mgcv::gam(formula,
                     method = method,
                     data = df_fit)
  
    nx_pred <- 1000
    xind_pred <- lapply(1:K,
                        function(x) {
                        rn_x <- range(X[, x])
                        seq(rn_x[1], rn_x[2], len = nx_pred)
                      })
    ## Sat Mar 27 20:07:40 2021 ------------------------------
    ## this part is removed because the graphes do not work
    # par(mfrow = c(2, 2))
    # for (p in 1:K) {
    #   plot(fit,
    #        select = p,
    #        residuals = TRUE,
    #        shade = TRUE)
    #   lines(xind_pred[[p]],
    #         f_list[[p]](xind_pred[[p]]),
    #         col = "red",
    #         lwd = 2,
    #         lty = 2)
    # }
  } else {
    print("number of functions is not consistent")
    break
  }
  augment <- broom::augment(fit) %>%
    mutate(bias = abs(.resid),
           # var = var(.resid),
           ## mse is the bias^2 + variance
           mse = .resid^2 + .se.fit^2,
           cover = case_when((y >= .fitted - 1.96 * .se.fit) & 
                               (y <= .fitted + 1.96 * .se.fit) ~ 1,
                             TRUE ~ 0))
  ## Sun Mar 28 09:08:44 2021 ------------------------------
  ## add bias mse and coverage
  return(list(mean(augment$bias),
              mean(augment$mse),
              mean(augment$cover),
              # broom::tidy(fit), 
              # broom::glance(fit),
              ## save augment here just in case 
              augment))
}
```

```{r eval=FALSE, include=TRUE}
# f1 <- function(x) sin(pi * x) / (x + 1)
# f2 <- function(x) sin(pi * x / 2) * x
# f3 <- function(x) sin(pi * x / 4)
# f4 <- function(x) cos(pi * x)
# f5 <- function(x) cos(pi * x / 2) * x^2
# f6 <- function(x) sin(pi * x) * x + cos(pi * x) * x
# 
# sim3p <- sim2p <- sim4p <- list()
# sim3u <- sim2u <- sim4u <- list()
# for (i in 1:1000) {
#   sim3p[[i]] <- purrr::map2(.x = rep(c(200, 300, 500), times = 3),
#                .y = list(list(f1, f3, f4),
#                          list(f1, f3, f5),
#                          list(f2, f3, f4),
#                          list(f1, f3, f4),
#                          list(f1, f3, f5),
#                          list(f2, f3, f4),
#                          list(f1, f3, f4),
#                          list(f1, f3, f5),
#                          list(f2, f3, f4)),
#                ~ simulate_gam(N = .x, K = 3,
#                               f_list = .y)) %>%
#               array()
#   sim3u[[i]] <- purrr::map2(.x = rep(c(200, 300, 500), times = 3),
#                .y = list(list(f1, f3, f4),
#                          list(f1, f3, f5),
#                          list(f2, f3, f4),
#                          list(f1, f3, f4),
#                          list(f1, f3, f5),
#                          list(f2, f3, f4),
#                          list(f1, f3, f4),
#                          list(f1, f3, f5),
#                          list(f2, f3, f4)),
#                ~ simulate_gam(N = .x, K = 3,
#                               f_list = .y,
#                               method = "REML")) %>%
#               array()
#   # Sat Mar 27 20:49:18 2021 ------------------------------
#   sim2u[[i]] <- purrr::map2(.x = rep(c(200, 300, 500), times = 3),
#                  .y = list(list(f1, f2),
#                            list(f4, f3),
#                            list(f5, f1),
#                            list(f1, f2),
#                            list(f4, f3),
#                            list(f5, f1),
#                            list(f1, f2),
#                            list(f4, f3),
#                            list(f5, f1)),
#                  ~ simulate_gam(N = .x, K = 2,
#                                 f_list = .y,
#                                 method = "REML")) %>%
#                 array()
#   sim2p[[i]] <- purrr::map2(.x = rep(c(200, 300, 500), times = 3),
#                  .y = list(list(f1, f2),
#                            list(f4, f3),
#                            list(f5, f1),
#                            list(f1, f2),
#                            list(f4, f3),
#                            list(f5, f1),
#                            list(f1, f2),
#                            list(f4, f3),
#                            list(f5, f1)),
#                  ~ simulate_gam(N = .x, K = 2,
#                                 f_list = .y)) %>%
#                 array()
#   # Sat Mar 27 20:49:30 2021 ------------------------------
#   sim4p[[i]] <- purrr::map2(.x = rep(c(200, 300, 500), times = 3),
#                  .y = list(list(f1, f2, f4, f3),
#                            list(f4, f3, f5, f6),
#                            list(f5, f1, f2, f6),
#                            list(f1, f2, f4, f3),
#                            list(f4, f3, f5, f6),
#                            list(f5, f1, f2, f6),
#                            list(f1, f2, f4, f3),
#                            list(f4, f3, f5, f6),
#                            list(f5, f1, f2, f6)),
#                  ~ simulate_gam(N = .x, K = 4,
#                                 f_list = .y)) %>%
#                 array()
#   sim4u[[i]] <- purrr::map2(.x = rep(c(200, 300, 500), times = 3),
#                  .y = list(list(f1, f2, f4, f3),
#                            list(f4, f3, f5, f6),
#                            list(f5, f1, f2, f6),
#                            list(f1, f2, f4, f3),
#                            list(f4, f3, f5, f6),
#                            list(f5, f1, f2, f6),
#                            list(f1, f2, f4, f3),
#                            list(f4, f3, f5, f6),
#                            list(f5, f1, f2, f6)),
#                  ~ simulate_gam(N = .x, K = 4,
#                                 f_list = .y,
#                                 method = "REML")) %>%
#                 array()
#   cat("this is the ", i, "iteration \n")
# }
# save(sim2p, file = "simulation_2function_penalized.Rdata")
# save(sim3p, file = "simulation_3function_penalized.Rdata")
# save(sim4p, file = "simulation_4function_penalized.Rdata")
# save(sim2u, file = "simulation_2function_unpenalized.Rdata")
# save(sim3u, file = "simulation_3function_unpenalized.Rdata")
# save(sim4u, file = "simulation_4function_unpenalized.Rdata")
```

```{r}
load("simulation_2function_penalized.Rdata")
load("simulation_3function_penalized.Rdata")
load("simulation_4function_penalized.Rdata")
load("simulation_2function_unpenalized.Rdata")
load("simulation_3function_unpenalized.Rdata")
load("simulation_4function_unpenalized.Rdata")

```

```{r}
sim2p_bias <- sim3p_bias <- sim4p_bias <-
  sim2u_bias <- sim3u_bias <- sim4u_bias <- data.frame()
sim2p_mse <- sim3p_mse <- sim4p_mse <-
  sim2u_mse <- sim3u_mse <- sim4u_mse <- data.frame()
sim2p_cover <- sim3p_cover <- sim4p_cover <-
  sim2u_cover <- sim3u_cover <- sim4u_cover <- data.frame()
for (j in seq_along(1:9)){
  for (i in seq_along(1:1000)) {
    sim2p_bias[j, i] <- sim2p[[i]][[j]][1]
    sim3p_bias[j, i] <- sim3p[[i]][[j]][1]
    sim4p_bias[j, i] <- sim4p[[i]][[j]][1]
    sim2u_bias[j, i] <- sim2u[[i]][[j]][1]
    sim3u_bias[j, i] <- sim3u[[i]][[j]][1]
    sim4u_bias[j, i] <- sim4u[[i]][[j]][1]

    sim2p_mse[j, i] <- sim2p[[i]][[j]][2]
    sim3p_mse[j, i] <- sim3p[[i]][[j]][2]
    sim4p_mse[j, i] <- sim4p[[i]][[j]][2]
    sim2u_mse[j, i] <- sim2u[[i]][[j]][2]
    sim3u_mse[j, i] <- sim3u[[i]][[j]][2]
    sim4u_mse[j, i] <- sim4u[[i]][[j]][2]

    sim2p_cover[j, i] <- sim2p[[i]][[j]][3]
    sim3p_cover[j, i] <- sim3p[[i]][[j]][3]
    sim4p_cover[j, i] <- sim4p[[i]][[j]][3]
    sim2u_cover[j, i] <- sim2u[[i]][[j]][3]
    sim3u_cover[j, i] <- sim3u[[i]][[j]][3]
    sim4u_cover[j, i] <- sim4u[[i]][[j]][3]
  }
}

mse <- 
  rbind(sim2p_mse, sim3p_mse, sim4p_mse,
        sim2u_mse, sim3u_mse, sim4u_mse) %>%
  transmute(mse = rowMeans(.)) 
bias <- 
  rbind(sim2p_bias, sim3p_bias, sim4p_bias,
        sim2u_bias, sim3u_bias, sim4u_bias) %>%
  transmute(bias = rowMeans(.)) 
cover <- 
  rbind(sim2p_cover, sim3p_cover, sim4p_cover,
        sim2u_cover, sim3u_cover, sim4u_cover) %>%
  transmute(cover = rowMeans(.)) 
```

```{r}
result <-  cbind(mse, bias, cover) %>%
  mutate(`#function` = rep(c("two", "three", "four", 
                             "two", "three", "four"), each = 9),
         functions = c(rep(c("f12", "f34", "f51"), 3),
                       rep(c("f134", "f135", "f234"), 3),
                       rep(c("f1243", "f4356", "f5126"), 3),
                       rep(c("f12", "f34", "f51"), 3),
                       rep(c("f134", "f135", "f234"), 3),
                       rep(c("f1243", "f4356", "f5126"), 3)),
         penalty = rep(c("penalized", "unpenalized"), each = 27),
         simulation = rep(rep(c(200, 300, 500), each = 3), time = 6))

result %>% 
  filter(simulation == 500, 
         `#function` == "four") 

result %>% 
  filter(functions == "f4356") 

result %>% 
  filter(penalty == "penalized", 
         simulation == 300)

result %>% 
  filter(penalty == "unpenalized", 
         simulation == 300)
```

The complexity of functions order from easy to hard: f3&4, f1&2, f5&6.
In conclusion, under the control of simulation times and number of functions, 
the penalized estimation has smaller mse, smaller bias, and larger 95% CI coverage,
regarding as a better performance comparing to than unpenalized methods; 
different penalized strategies share the similar behaviors on the performance:
with more complicated functions, the model suffers smaller coverage,
large bias, and large mean square error, which indicate the possibility of overfitting.

Controlling for the complexity (also the number of function), 
the penalized performs still better than the unpenalized with smaller mse, smaller bias,
and larger 95% CI coverage. The different penality strategies behave differently 
increasing simulations: for the penalized, more simulation shows better performance of the model,
the 95% coverage increased, and both bias and mse decrease at certain level; 
for the unpenlized, the more simulation indicates worse performance,
the coverage shrinks with larger simulation number, also with larger mse and bias.

For models with the same simulation number, the more complicated the functions decides 
the status of underfitting or overfitting; hence the mse and bias varied case by case.
In general, the more functions included in the model the better the performance.

# question6

$$
\pmb {\Phi_{N \times K}} = [\pmb c_{N \times 1}, \pmb {\Phi^{'}_{N \times (K-1)}}]
\\
\phi_i=[\phi_{i1}, \ ...\ \phi_{iK}]
\\
\phi_{ij}^{*}(x) = \phi_{ij}(x) - \frac {1} {N} \sum_{i=1}^N \phi_{ij} (x);
\\
\phi_{i1}^{*}(x) = \phi_{i1}(x) - \frac {1} {N} \sum_{i=1}^N \phi_{i1} (x) 
= c - \frac {1} {N}\sum_{i=1}^N c = 0
$$
Because the constant term is subtracted from the matrix, there is only one column of zeros.
the new basis matrix $\mathcal \Phi^*$ does not contain the constant vector c in the column space, and also less than full rank.


$$
f^*(x) = \pmb{\Phi^*\xi} = \pmb{\Phi\xi} - \pmb {1\ 1^{\top} \Phi \xi}/N = f(x) - \pmb {1}c
$$

more generally, we can define $1^{\top} \Phi \xi / N = c$
after pivoting with the constant c to be the first column 
and then subtraction on the constant vector,
then the new centralized matrix does not contain the original c anymore.
Moreover, because the first column is all zeros, 
the new basis matrix $\mathcal \Phi^*$ is less than full rank.

$$
\pmb {\Phi^*_{N \times K}} = [\pmb {0_{N \times 1}, \ \Phi^{*'}_{N \times (K-1)}}]
$$
The new column mean centered $\mathcal \Phi^*$ does not contain the original constant c vector.
In practice, we restrict the $1^{\top} \Phi \xi / N = c = 0$ which makes the process much easier.
We manually 
$$
\mathcal {f}(\pmb x) = \pmb{\Phi \xi}
\\
\sum_{i=1} ^N f(x_i) = \pmb {1 ^{\top} \Phi \xi} = 0, \text{ for any } \pmb \xi 
\\
\text{then, } \pmb {1^{\top} \Phi} = 0
\\
\pmb \Phi ^{*} = \pmb{\Phi} - \pmb {1\ 1^{\top} \Phi}/N
$$



