---
title: "01_fisrt_step"
author: "Randy"
date: "1/12/2021"
output: html_document
---
# 1. First Step in the analysis of functional data

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo  =  TRUE)
library(fda)
```

## 1.1 Basis expansions
```{r}
spline.basis <-
    create.bspline.basis(rangeval = c(0,10), 
                         nbasis = 5)
plot(spline.basis, lty = 1, lwd = 2)

?create.bspline.basis
## The B-spline basis is used for non-periodic functions. 
## B-spline basis functions are polynomial segments jointed
## end-to-end at at argument values called knots, 
## breaks or join points. 
## 
## The segments have specifiable smoothness across these breaks.
## B-spline basis functions have the advantages 
## of very fast computation and great flexibility. 

fourier.basis <-
    create.fourier.basis(rangeval = c(0,10), 
                         nbasis = 5)
plot(fourier.basis, lty = 1, lwd = 2)
## the Fourier basis is the constant function,
## then there are sine and cosine functions of 
## period equal to the length of the interval.

monomial.basis <- 
    create.monomial.basis(rangeval = c(0,10),
                          nbasis = 5)
plot(monomial.basis, lty = 1, lwd = 2)
```


```{r}
## the function smooth.basis is used 
## to create a functional data object Wiener.fd which
## contains the expansion coefficients cnm 
## as well as information about the basis
## functions used to expand the random walk function. 

Wiener <- cumsum(rnorm(10000))/100 
## random walk on [0,K], K = 10ˆ4
plot.ts(Wiener, xlab = "", ylab = "")

B25.basis <- 
    create.bspline.basis(rangeval = c(0,10000),
                         nbasis = 25)
Wiener.fd <- 
    smooth.basis(y = Wiener, 
                 fdParobj = B25.basis)

lines(Wiener.fd, lwd = 3, col = "red")
```

## 1.2 Sample mean and covariance

```{r}
N = 50
W.mat <- matrix(0, ncol = N, nrow = 10000)

for(n in 1:N){
    W.mat[, n] = cumsum(rnorm(10000))/100
}

B25.basis <- 
    create.bspline.basis(rangeval = c(0,10000), 
                         nbasis = 25)
W.fd <- smooth.basis(y = W.mat, 
                     fdParobj = B25.basis)
plot(W.fd, col = "gray", lty = 1, ylab = "", xlab = "")

W.mean <- mean(W.fd$fd)
W.sd <- std.fd(W.fd$fd)

lines(W.sd, lwd = 3)
lines(W.mean, lty = 2, lwd = 3)
```


```{r}
#Use the object W.fd generated in the previous example.
str(W.fd)

W.cov = var.fd(W.fd$fd) 
# $fd extracts function values 
grid = (1:100) * 100

##Values a Two-argument Functional Data Object
# A vector of argument values 
# for the first argument s of 
# the functional data object to be evaluated
?eval.bifd

W.cov.mat = eval.bifd(sevalarg = grid,
                      tevalarg = grid, 
                      bifd = W.cov)
## draws perspective plots of a surface 
## over the x–y plane. 
## persp is a generic function.
persp(x = grid, 
      y = grid, 
      ## z is a matrix containing the values 
      ## to be plotted (NAs are allowed)
      z = W.cov.mat, 
      xlab = "s",
      ylab = "t", 
      zlab = "c(s, t)")
## contour plot, or add contour lines
##  to an existing plot
contour(x = grid, 
        y = grid, 
        z = W.cov.mat, 
        lwd = 2)
```

## 1.3 Principal component functions
estimated functional principal compnents
EFPC related to covariance function

```{r}
## Functional Principal components analysis aims 
## to display types of variation 
## across a sample of functions. 
## 
## Principal components analysis is 
## an exploratory data analysis 
## that tends to be an early part of many projects. 
## These modes of variation are called $principal components$ 
## or $harmonics.$ 
## 
## This function computes these harmonics,
##  the eigenvalues that indicate 
##  how important each mode of variation, 
##  and harmonic scores for individual functions.
## 
## If the functions are multivariate, 
## these harmonics are combined into 
## a composite function that summarizes 
## joint variation among the several functions 
## that make up a multivariate functional observation.
W.pca1 = pca.fd(W.fd$fd, 
               ## the number of harmonics or 
               ## principal components to compute
               nharm = 3)
W.pca2 = pca.fd(W.fd$fd, 
               ## the number of harmonics or 
               ## principal components to compute
               nharm = 5)

View(W.pca1)
str(W.pca1)

plot(W.pca1$harmonics, lwd = 3)
plot(W.pca2$harmonics, lwd = 5)
W.pca1$varprop
W.pca2$varprop
```

## 1.4 Example: Analysis of BOA stock return

```{r}
data(package = "refund")


BOA <- read.table("BOA.txt", header = TRUE)
Dates <- dimnames(BOA)[[1]]
BOA <- data.matrix(BOA)

Outlier <- which(Dates == "08/26/2004")
BOA <- BOA[-Outlier, ]
N <- dim(BOA)[1]
M <- dim(BOA)[2]

Times <- seq(0, 6.5, length = M)
log_BOA <-  log(BOA) - matrix(log(BOA)[, 1],
                              nrow = N,
                              ncol = M)
bspline_basis <- 
    create.bspline.basis(rangeval = c(0,6.5),
                         norder = 4,
                         nbasis = 200)
log_BOA_f <-  
    Data2fd(Times,
            t(log_BOA),
            basisobj = bspline_basis)

plot(log_BOA_f[1:10],
     xlab = "",
     ylab = "",
     lwd = 1.5)
```


```{r}
muhat<-mean.fd(log_BOA_f)
sdhat<-sd.fd(log_BOA_f)

SE_hat_U<-fd(basisobj=bspline_basis) 
# create upper CI bound
SE_hat_L<-fd(basisobj=bspline_basis) 
# create lower CI bound

SE_hat_U$coefs<-2*sdhat$coefs/sqrt(N) + muhat$coefs
SE_hat_L$coefs<- -2*sdhat$coefs/sqrt(N) + muhat$coefs
plot.fd(SE_hat_U,ylim=c(-0.002,0.002),
        col='red',lty=2,
        xlab="",ylab="")
plot.fd(SE_hat_L,add=TRUE,
        col='red',lty=2)
plot.fd(muhat,add=TRUE)
```

One of the fundamental assumptions in many
financial tools is that the log returns 
can be modeled using Brownian motion
(or some variation of it), 
which agrees with these results.

## 1.5 Example: DTI Diffusoin tensor imaging
```{r}
library(refund)
data(DTI)
Y <- DTI$cca
Y <- Y[-c(126, 130, 131, 125, 319, 321), ] # missing values
N <- dim(Y)[1]
M <- dim(Y)[2]
argvals <- seq(0, 1, length = M)
data_basis <- create.bspline.basis(c(0, 1), nbasis = 10)
Y.f <- Data2fd(argvals, t(Y), data_basis)
dev.new(width = 8, height = 6)
graphics::plot(Y.f,
  col = 1,
  type = "l",
  lty = 1,
  xlab = "", ylab = "",
  ylim = c(0.1, .9)
)

lines(mean(Y.f), lwd = 2)
lines(mean(Y.f) + std.fd(Y.f), lwd = 2, lty = 2, col = "green")
lines(mean(Y.f) + 2 * std.fd(Y.f), lwd = 2, lty = 2, col = "yellow")
lines(mean(Y.f) + 3 * std.fd(Y.f), lwd = 2, lty = 2, col = "red")
lines(mean(Y.f) - std.fd(Y.f), lwd = 2, lty = 2, col = "green")
lines(mean(Y.f) - 2 * std.fd(Y.f), lwd = 2, lty = 2, col = "yellow")
lines(mean(Y.f) - 3 * std.fd(Y.f), lwd = 2, lty = 2, col = "red")
```

