---
title: "02_exploratory"
author: "Randy"
date: "1/12/2021"
output: html_document
---
# 2. Further topics in exploratory ananlysis of the functional data
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 2.1 Derivatives

## 2.2 Penalized smoothing

which imposes a penalty on functions that are too \wiggly."

Using sufficiently large M changes the curves somewhat, 
but they essentially look the same as the original data curves.

1) GCV is invariant to rotations of
the original regression problem 
2) variance of feature vectors is “spread out”
evenly across the transformed matrix X′

```{r}
# set up a saturated basis: as many basis
# functions as observations
nbasis = 365
yearRng = c(0,365)
daybasis = create.fourier.basis(yearRng, nbasis)
logprecav = CanadianWeather$dailyAv[, , 'log10precip']
dayprecfd <- with(CanadianWeather, smooth.basis(day.5,
logprecav, daybasis,
fdnames=list("Day", "Station", "log10(mm)"))$fd )
for(i in 1:5){
    plot(logprecav[,i],
         axes=FALSE,
         xlab="Day",
         ylab="log10(mm)",
         main=CanadianWeather$place[i])
    lines(dayprecfd[i],col=2)
    axisIntervals(1)
    axis(2)
    readline("Press <return to continue>")
}
```

```{r}
Lcoef = c(0,(2*pi/diff(yearRng))^2,0,1)
harmaccelLfd = vec2Lfd(Lcoef, yearRng)

loglam = 4:9
nlam = length(loglam)
dfsave = rep(NA,nlam)
names(dfsave) = loglam
gcvsave = dfsave

for (ilam in 1:nlam) {
    cat(paste('log10 lambda =', loglam[ilam], '\n'))
    lambda = 10^loglam[ilam]
    fdParobj = fdPar(daybasis, harmaccelLfd, lambda)
    smoothlist = smooth.basis(day.5, 
                              logprecav,
                              fdParobj)
    dfsave[ilam] = smoothlist$df
    gcvsave[ilam] = sum(smoothlist$gcv)
}

plot(loglam, gcvsave, type='b', lwd=2)
```


```{r}
lambda = 1e6
fdParobj = fdPar(daybasis, 
                 harmaccelLfd, 
                 lambda)
logprec.fit = smooth.basis(day.5, 
                           logprecav, 
                           fdParobj)
logprec.fd = logprec.fit$fd
fdnames = list("Day (July 1 to June 30)", 
                "Weather Station" = CanadianWeather$place, 
                "Log 10 Precipitation (mm)")
logprec.fd$fdnames = fdnames


# plot smoothed curves for the first five locations
for(i in 1:5){
    plot(logprecav[,i],
         axes=FALSE,
         xlab="Day",
         ylab="log10(mm)",
         main=CanadianWeather$place[i])
    lines(dayprecfd[i],col=2)
    axisIntervals(1)
    axis(2)
    lines(logprec.fd[i], lwd=3)
    readline("Press <return to continue>")
}

```


## 2.3 Curve alignment

```{r}
age=growth$age
heightBasis=create.bspline.basis( c(1,18), 35, 6, age)
heightPar=fdPar(heightBasis, 3, 10^(-0.5))
heightSmooth=smooth.basis(age, growth$hgtf, heightPar)
plot(heightSmooth, lwd=2, xlab="age", ylab="height (cm)")
```


```{r}
accelUnreg = deriv.fd(heightSmooth$fd, 2)
plot(accelUnreg[,1], lwd=2, 
     xlab="Age", 
     ylab="Acceleration",
     ylim=c(-4,2))
mean.accelUnreg=mean(accelUnreg)
lines(mean.accelUnreg, lwd=4, col="black")
```


```{r}
regList=register.fd(yfd=accelUnreg)
accelReg=regList$regfd
# $regfd extracts the registered functions
plot(accelReg, xlab="Age", 
     ylab="Acceleration",
     ylim=c(-4,3))
```


```{r}
warpFunctions=regList$warpfd
# $warpfd extracts the warping functions
APList=AmpPhaseDecomp(xfd=accelUnreg, 
                      yfd=accelReg, 
                      hfd=warpFunctions)

str(APList)
APList$RSQR
# $RSQR extract proportion of 
# the total variation 
# due to phase variation
APList$MS.amp
APList$MS.pha 
```






