---
title: "Explorations and Experiments on INLA with NHANES data, PartII"
subtitle: "BIOS 7720"
author: "Randy"
date: April 20, 2021
header-includes:
   - \usepackage{bm}
output:
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

## Random Walk and Smoothing functions

$$
\int \Big(f^{(m)}(x) \Big)^2dx\ \approx\ d^{-(2m-1)} \sum_{i=m+1}^n[\nabla^m f(x_i)] ^2
$$

$d$ is the samll and equal space for $d_i = x_i - x_{i-1}$;    
$m^{th}$ derivative of $f$ is continuous;    
$\nabla^m$ is the $m^th$ order backward difference operator:        
$\nabla^1 f(x) = f(x_i) -f(x_{i-1})$        
$\nabla^2 f(x) = f(x_i) -2f(x_{i-1}) + f(x_{i-2})$        


## Random Walk and Smoothing functions

$$
\nabla^m f(x_i)\ \overset{iid}{\sim}\ Normal(0,\ \sigma^2_f\ ),\ i = m+1,\ ...,\ n.
$$

$$
\sqrt {\frac {\| \mathcal{Q_m} \|_+} {2\pi\sigma_f^2}} exp \Bigg(-\frac {1}{2\sigma^2_f} f^{\top}\mathcal{Q_m}f\Bigg)
$$

## simulation and comparison

```{r}
library(INLA)
library(brinla)
library(tidyverse)
library(mgcv)

set.seed(555)
n <- 100
x <- seq(0, 1, , n)
f.true <- (sin(2 * pi * x^3))^3
y <- f.true + rnorm(n, sd = 0.2)
```


## RW1 and RW2
```{r}
data.inla <- data.frame(y = y, x = x)
formula1 <- y ~ -1 + f(x, model = "rw1", constr = FALSE)
system.time(result1 <- inla(formula1, data = data.inla))

formula2 <- y ~ -1 + f(x, model = "rw2", constr = FALSE)
system.time(result2 <- inla(formula2, data = data.inla))
# names(inla.models()$latent)
```


## RW1 and RW2

```{r}
head(result1$summary.random$x) %>% round(4)
head(result2$summary.random$x) %>% round(4)
```

## cubic

```{r}
#### smooth_spline() ----------
fit.ss <- smooth.spline(x, y)
res <- (fit.ss$yin - fit.ss$y) / (1 - fit.ss$lev)
fhat3 <- fit.ss$y 
## lower bound
f.lb3 <- fhat3 - 2 * sd(res) * sqrt(fit.ss$lev) 
## upper bound
f.ub3 <- fhat3 + 2 * sd(res) * sqrt(fit.ss$lev)
```

## mgcv gam

```{r}
#### gam() Sun Apr 04 09:19:51 2021 ---------
system.time(fit.gam <- gam(y ~ s(x)))
res.gam <- predict(fit.gam, se.fit = TRUE)
## fitted curve
fhat4 <- res.gam$fit 
## lower bound
f.lb4 <- res.gam$fit - 2 * res.gam$se.fit 
## upper bound
f.ub4 <- res.gam$fit + 2 * res.gam$se.fit 
```

```{r include=FALSE}
fhat1 <- result1$summary.random$x$mean
f.lb1 <- result1$summary.random$x$"0.025quant"
f.ub1 <- result1$summary.random$x$"0.975quant"
data.plot1 <- data.frame(y = y, x = x, 
                        f.true = f.true, 
                        fhat = fhat1, 
                        f.lb = f.lb1, 
                        f.ub = f.ub1)


fhat2 <- result2$summary.random$x$mean
f.lb2 <- result2$summary.random$x$"0.025quant"
f.ub2 <- result2$summary.random$x$"0.975quant"
data.plot2 <- data.frame(y = y, x = x, 
                        f.true = f.true, 
                        fhat = fhat2, 
                        f.lb = f.lb2, 
                        f.ub = f.ub2)


data.plot3 <- data.frame(y = y, x = x, 
                        f.true = f.true, 
                        fhat = fhat3, 
                        f.lb = f.lb3, 
                        f.ub = f.ub3)

data.plot4 <- data.frame(y = y, x = x, 
                        f.true = f.true, 
                        fhat = fhat4, 
                        f.lb = f.lb4, 
                        f.ub = f.ub4)

plot1 <- data.plot1 %>%
  ggplot(aes(x = x, y = y)) +
  geom_line(aes(y = fhat), size = 2, color = "darkgreen") +
  geom_line(aes(y = f.true), size = 2, color = "indianred") +
  geom_ribbon(aes(ymin = f.lb, ymax = f.ub), alpha = 0.2) +
  geom_point(aes(y = y)) +
  theme_bw(base_size = 20) +
  ggtitle("RW1")

plot2 <- data.plot2 %>%
  ggplot(aes(x = x, y = y)) +
  geom_line(aes(y = fhat), size = 2, color = "darkgreen") +
  geom_line(aes(y = f.true), size = 2, color = "indianred") +
  geom_ribbon(aes(ymin = f.lb, ymax = f.ub), alpha = 0.2) +
  geom_point(aes(y = y)) +
  theme_bw(base_size = 20) +
  ggtitle("RW2")

plot3 <- data.plot3 %>%
  ggplot(aes(x = x, y = y)) +
  geom_line(aes(y = fhat), size = 2, color = "darkgreen") +
  geom_line(aes(y = f.true), size = 2, color = "indianred") +
  geom_ribbon(aes(ymin = f.lb, ymax = f.ub), alpha = 0.2) +
  geom_point(aes(y = y)) +
  theme_bw(base_size = 20) +
  ggtitle("cubic")

plot4 <- data.plot4 %>%
  ggplot(aes(x = x, y = y)) +
  geom_line(aes(y = fhat), size = 2, color = "darkgreen") +
  geom_line(aes(y = f.true), size = 2, color = "indianred") +
  geom_ribbon(aes(ymin = f.lb, ymax = f.ub), alpha = 0.2) +
  geom_point(aes(y = y)) +
  theme_bw(base_size = 20) +
  ggtitle("mgcv")
```


```{r fig.height=10, fig.width=10}
gridExtra::grid.arrange(plot1, plot2, plot3, plot4, nrow = 2)
```

## the cherry on the top - prior

* To allow the data to speak for themselves,          
weakly informative or non-informative priors             
are often used in this situation. 
```{r}
a1 <- 3
b1 <- 5e-5
a2 <- -0.5
b2 <- 5e-5

lgprior1 <- list(prec = list(param = c(a1, b1)))
lgprior2 <- list(prec = list(param = c(a2, b2)))

formula5 <- y ~ -1 + f(x, model = "rw2", 
                      constr = FALSE, 
                      hyper = lgprior2)

system.time(
result5 <- inla(formula5, 
               data = data.inla, 
               ## used to specify the prior on delta
               ## and hyper in f() in f() on tao
               control.family = list(hyper = lgprior1))
)
```

## the cherry on the top - prior

* the resulting posterior distributions do not differ much                
from those given by the default priors

```{r}
## posterior mean
fhat <- result5$summary.random$x$mean 
## 2.5% percentile
f.lb <- result5$summary.random$x$"0.025quant"
## 97.5% percentile
f.ub <- result5$summary.random$x$"0.975quant" 


data.plot5 <- data.frame(y = y, x = x, 
                        f.true = f.true, 
                        fhat = fhat, 
                        f.lb = f.lb, 
                        f.ub = f.ub)

plot5 <- data.plot5 %>%
  ggplot(aes(x = x, y = y)) +
  geom_line(aes(y = fhat), size = 2, color = "darkgreen") +
  geom_line(aes(y = f.true), size = 2, color = "indianred") +
  geom_ribbon(aes(ymin = f.lb, ymax = f.ub), alpha = 0.2) +
  geom_point(aes(y = y)) +
  theme_bw(base_size = 20) +
  ggtitle("prior")
```


```{r fig.height=5, fig.width=15}
gridExtra::grid.arrange(plot2, plot4, plot5, nrow = 1)
```










